---
output: github_document
---

# Rmonize 2.0.0 (release : 2025-06-26)

> Attention: Some changes to functions in the current version of madshapR may require
updates of existing code.

## Superseded object.

| previous version (1.1.0 and older) | version 2.0.0          |
|------------------------------------|------------------------|
| Rmonize_DEMO                       | Rmonize_examples       |

## Superseded parameters.

In functions show_harmo_error(), harmonized_dossier_evaluate(), 
harmonized_dossier_summarize() and harmonized_dossier_visualize(), 
the parameters have been simplified into one and only "dossier"
https://github.com/maelstrom-research/Rmonize/issues/110
https://github.com/maelstrom-research/Rmonize/issues/109
https://github.com/maelstrom-research/Rmonize/issues/108
https://github.com/maelstrom-research/Rmonize/issues/98
https://github.com/maelstrom-research/Rmonize/issues/93
https://github.com/maelstrom-research/Rmonize/issues/92

previous version (1.1.0 and older)

```{r, eval=FALSE}
harmonized_dossier_evaluate(
  harmonized_dossier,dataschema,taxonomy,as_dataschema_mlstr)

harmonized_dossier_summarize(
  harmonized_dossier,group_by,dataschema,data_proc_elem,
  taxonomy,valueType_guess)

harmonized_dossier_visualize(
  harmonized_dossier,bookdown_path,group_by,harmonized_dossier_summary,
  dataschema,data_proc_elem,valueType_guess,taxonomy)

```
 
 version 2.0.0

```{r, eval=FALSE}

harmonized_dossier_evaluate(harmonized_dossier)
harmonized_dossier_summarize(harmonized_dossier)
harmonized_dossier_visualize(harmonized_dossier,bookdown_path)

```



## Superseded function behaviors and/or output structures.

In `harmonized_dossier_evaluate()`, the columns generated in the outputs have 
been renamed as follows :

| previous version (1.1.0 and older) | current version (2.0.0)         |
|------------------------------------|---------------------------------|
| index                              | Index                           |
| name                               | Variable name                   |
| label                              | Variable label                  |
| valueType                          | Data dictionary valueType       | 
| Categories::label                  | Categories in data dictionary   | 
| Categories::missing                | Non-valid categories            |

In `harmonized_dossier_summarize()`, the columns generated in the outputs have 
been renamed as follows :

| previous version (1.1.0 and older) | current version (2.0.0)         |
|------------------------------------|---------------------------------|
| index in data dict.name            | Index                      | 
| name                               | Variable name              |
| label                              | Variable label             |
| Estimated dataset valueType        | Suggested valueType        | 
| Actual dataset valueType           | Dataset valueType          |
| Total number of observations       | Number of rows             |
| Nb. distinct values                | Number of distinct values  |
| Nb. valid values                   | Number of valid values     |
| Nb. non-valid values               | Number of non-valid values |
| Nb. NA                             | Number of empty values     |
| % total Valid values               | % Valid values             |
| % Non-valid values                 | % Non-valid values         |
| % NA                               | % Empty values             |
|------------------------------------|---------------------------------|


## Bug fixes and improvements

## Enhancements in the assessment, the summary and the visual reports!

* The assessment and summary reports had some updates, such as renamed columns 
and bug corrections.
https://github.com/maelstrom-research/Rmonize/issues/104
https://github.com/maelstrom-research/Rmonize/issues/103
https://github.com/maelstrom-research/Rmonize/issues/89
https://github.com/maelstrom-research/Rmonize/issues/88
https://github.com/maelstrom-research/Rmonize/issues/87
https://github.com/maelstrom-research/Rmonize/issues/86
https://github.com/maelstrom-research/Rmonize/issues/85
https://github.com/maelstrom-research/Rmonize/issues/84
https://github.com/maelstrom-research/Rmonize/issues/68
https://github.com/maelstrom-research/Rmonize/issues/21

* The visual reports have been improved, including better visual outputs and
color palettes, and new features such as total number of rows next to the bar charts.

https://github.com/maelstrom-research/Rmonize/issues/57
https://github.com/maelstrom-research/Rmonize/issues/53
https://github.com/maelstrom-research/Rmonize/issues/49
https://github.com/maelstrom-research/Rmonize/issues/48
https://github.com/maelstrom-research/Rmonize/issues/39
https://github.com/maelstrom-research/Rmonize/issues/37
https://github.com/maelstrom-research/Rmonize/issues/33
https://github.com/maelstrom-research/Rmonize/issues/32
https://github.com/maelstrom-research/Rmonize/issues/29



# Rmonize 1.1.0

## Bug fixes and improvements

* To process the data during testing, the DataSchema and/or the 
Data Processing Elements and/or input datasets might not be available. 
To be able to perform testings on harmonization, an additional 
parameter `.debug` has been added
https://github.com/maelstrom-research/Rmonize/issues/56

* The report function can now work when the code is indented in the 
Data Processing Elements.
https://github.com/maelstrom-research/Rmonize/issues/54

* The function `show_harmo_error()` now allows the user to avoid showing warnings
https://github.com/maelstrom-research/Rmonize/issues/52

## deprecated functions

To avoid confusion with help(function), the function `Rmonize_help()` has been 
renamed `Rmonize_website()`.

## Dependency changes

* set a minimum dplyr dependence to avoid bugs

# Rmonize 1.0.1

Bug corrections and enhancements after testing with real data.

## Bug fixes and improvements


### Improvement in handling pooled data

The functions `harmo_process()`, `pool_harmonized_dataset_create()`, 
`harmonized_dossier_create()`, `harmonized_dossier_evaluate()`, 
`harmonized_dossier_summarize()`, `harmonized_dossier_visualize()` share the 
same parameter "harmonized_col_dataset" which is (if exists) the name of the 
column referring the input dataset names. If this column exists and is declared 
by the user, this will be used across the pipeline as a grouping/separating 
variable. By default, the name of each dataset will be used instead.



rename DEMO_file_harmo into Rmonize_DEMO and update examples

suppress the parameter overwrite = TRUE in the functions xxx_visualize()

* https://github.com/maelstrom-research/Rmonize/issues/38

in visual reports, void confusing changes in color scheme in visual reports.
  

* https://github.com/maelstrom-research/Rmonize/issues/37

Histograms for date variables display valid ranges.
  

* https://github.com/maelstrom-research/Rmonize/issues/31

in reports, change % NA as proportion in reports.
  

* https://github.com/maelstrom-research/Rmonize/issues/29

`harmonized_dossier_visualize()` report shows variable labels in the same language.
  

* https://github.com/maelstrom-research/Rmonize/issues/28

put id_creation in script and in rule in dpe (as in direct_mapping)


* https://github.com/maelstrom-research/Rmonize/issues/27

Allow special characters in names of datasets and data_dicts


* https://github.com/maelstrom-research/Rmonize/issues/23

In visual reports, the bar plot only appears when there are multiple missing 
value types, otherwise only the pie chart is shown.


* https://github.com/maelstrom-research/Rmonize/issues/22

enhance harmonized_dossier_visualize() output


* https://github.com/maelstrom-research/Rmonize/issues/17

enhance `show_harmo_error()` output


* https://github.com/maelstrom-research/Rmonize/issues/5

in reports, all of the percentages are now included under "Other values 
(non categorical)", which gives a single value.


* https://github.com/maelstrom-research/Rmonize/issues/4

Function recode with special character is possible now

  
# Rmonize 1.0.0

Functions to support rigorous retrospective data harmonization processing, 
evaluation, and documentation across datasets in a dossier based on 
Maelstrom Research guidelines. The package includes the core functions to 
evaluate and format the main inputs that define the harmonization process, 
apply specified processing rules to generate harmonized data, diagnose 
processing errors, and summarize and evaluate harmonized outputs. 

This is still a work in progress, so please let us know if you used a function 
before and is not working any longer.

## Helper functions and objects

* `Rmonize_help()` Call the help center for full documentation
* `dowload_templates()` Call the help center to the download template page
* `Rmonize_DEMO` Built-in material allowing the user to test the package 
with demo data

## Assess and manipulate input files

* `as_data_proc_elem()` Validate and coerce any object as a Data Processing Elements
* `as_dataschema()`, `as_dataschema_mlstr()` Validate and coerce any object as 
the DataSchema
* `as_harmonized_dossier()` Validate and coerce any object as an harmonized dossier
* `dataschema_extract()` Extract and create the DataSchema from a data 
processing elements

## Data processing

* `harmo_process()` Generate harmonized dataset(s) and annotated 
Data Processing Elements. This function internally runs other functions, 
which are :

* `harmo_parse_process_rule()`,
`harmo_process_add_variable()`,`harmo_process_case_when()`,
`harmo_process_direct_mapping()`,`harmo_process_id_creation()`,
`harmo_process_impossible()`,`harmo_process_merge_variable()`,
`harmo_process_operation()`,`harmo_process_other()`,
`harmo_process_paste()`,`harmo_process_recode()`,
`harmo_process_rename()`,`harmo_process_undetermined()`
* `pooled_harmonized_dataset_create()` Generate the pooled dataset from 
harmonized datasets in a dossier

## Evaluation of the harmonization process

* `show_harmo_error()` Generate a summary of the annotated Data Processing Elements
* `data_proc_elem_evaluate()`,`dataschema_evaluate()`,
`harmonized_dossier_evaluate()`,`harmonized_dossier_summarize()`,
`harmonized_dossier_visualize()` Generate a quality assessment reports and 
summary statistics of inputs and outputs. 

## import from madshapR package:

* Shape and prepare input (datasets and data dictionaries) :

`as_data_dict()`,`is_data_dict()`,
`as_data_dict_mlstr()`,`is_data_dict_mlstr()`,
`as_dataset()`,`is_dataset()`,
`as_dossier()`,`is_dossier()`,
`as_taxonomy()`

* Extract and manipulate information from input :

`data_extract()`,`data_dict_extract()`,
`data_dict_apply()`,`dataset_zap_data_dict()`,`dossier_create()`
`valueType_adjust()`

* Assess input data :

`dataset_evaluate()`, `data_dict_evaluate()`,`dossier_evaluate()`,
`dataset_summarize()`,`dossier_summarize()`

* Visualize input data :

`bookdown_template()`,`bookdown_render()`,`bookdown_open()`,
`dataset_visualize()`


