---
title: "Example with Rmonize_DEMO"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example with Rmonize_DEMO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = T,
  results = "hide",
  eval = FALSE)

```

# Objective of the vignette

This vignette provides examples of applying Rmonize functions to prepare inputs, 
process data, and validate and consolidate harmonized data, using illustrative 
demo objects included with the package. The vignette focuses on demonstrating 
the usage of functions on inputs that already have valid structure and content. 
See the 
[Glossary](https://maelstrom-research.github.io/Rmonize-documentation/articles/a-Glossary-and-templates.html)
and 
[Reference](https://maelstrom-research.github.io/Rmonize-documentation/reference/index.html) 
pages for more details about the terms used in 
this document.


# Get started

## Install the package

```{r, eval=FALSE}
# To install Rmonize:
install.packages('Rmonize')

library(Rmonize)
# If you need help with the package, please use:
Rmonize_website()

# Downloadable templates are available here
Rmonize_templates()

# Demo files are available here, along with an online demonstration process 
Rmonize_DEMO

```

# Demo objects
Demo objects are available through the built-in 
[Rmonize_DEMO](https://maelstrom-research.github.io/Rmonize-documentation/reference/Rmonize_DEMO.html) object. They include example 
input datasets, input data dictionaries, DataSchema, Data Processing Elements 
(DPE), and harmonized datasets that provide illustrative examples of the 
structure and content of the main objects used by Rmonize functions.


```{r}
# To see contents
names(Rmonize_DEMO)
print(Rmonize_DEMO$dataset_TOKYO)	                     # An input dataset
print(Rmonize_DEMO$data_dict_TOKYO)             	     # An input data dictionary
print(Rmonize_DEMO$`data_processing_elements - final`) # A Data Processing Elements 
print(Rmonize_DEMO$`dataschema - final`)	             # A DataSchema

```


# Pipeline

## Prepare inputs

### DataSchema and Data Processing Elements (DPE)

The DataSchema and DPE are generally prepared from Excel 
[templates](https://maelstrom-research.github.io/Rmonize-documentation/articles/a-Glossary-and-templates.html) and imported into R. 
Separate documentation is provided for preparing Data Processing Elements. 
The DataSchema will be an R list with named elements Variables and Categories. 
The DPE will be a data frame. You can check the structure of each object and 
assign the correct attributes explicitly to ensure compatibility with Rmonize 
functions.

```{r}
# as_dataschema and as_data_proc_elem will check the structure of object and 
# assign attributes to them.

dataschema <- as_dataschema(Rmonize_DEMO$`dataschema - final`)
data_proc_elem <- as_data_proc_elem(Rmonize_DEMO$`data_processing_elements - final`)

```

<em>
Note: In the DEMO DPEs, all elements for three different input datasets are 
combined in one file. A DPE can also be prepared for each input dataset 
separately, and the individual DPEs combined as needed for processing. 
The DataSchema and DPE objects can be assigned explicitly
</em>

### Combine input datasets and data dictionaries in a dossier

If input data dictionaries (with metadata about variables and categories in 
input datasets) are available, they can be associated with their corresponding 
datasets using the function 
[data_dict_apply()](https://maelstrom-research.github.io/madshapR-documentation/reference/data_dict_apply.html). If not provided, minimal
data dictionaries will automatically be created to meet technical requirements 
of Rmonize functions as needed.


```{r}
# Associate metadata from input data dictionaries to the input datasets.

dataset_MELBOURNE <- data_dict_apply(
  dataset = Rmonize_DEMO$dataset_MELBOURNE,
  data_dict = Rmonize_DEMO$data_dict_MELBOURNE)

dataset_PARIS <- data_dict_apply(
  dataset = Rmonize_DEMO$dataset_PARIS,
  data_dict = Rmonize_DEMO$data_dict_PARIS)

dataset_TOKYO <- data_dict_apply(
  dataset = Rmonize_DEMO$dataset_TOKYO,
  data_dict = Rmonize_DEMO$data_dict_TOKYO)

```

For use in [harmo_process()](https://maelstrom-research.github.io/Rmonize-documentation/reference/harmo_process.html), one or more input 
datasets and any associated data dictionaries must be grouped into a named 
list (referred to as a "dossier" in Rmonize). This can be done explicitly with
[dossier_create()](https://maelstrom-research.github.io/madshapR-documentation/reference/dossier_create.html).

```{r}

# Group the datasets into a dossier object.
# NB: The names of the datasets in the dossier must match the column 
# input_dataset in the Data Processing Elements

dossier <- dossier_create( dataset_list = list(
  dataset_MELBOURNE, 
  dataset_PARIS, 
  dataset_TOKYO))

```

## Process data

When the input dossier, DataSchema, and DPE are prepared, you can initiate 
processing using [harmo_process()](https://maelstrom-research.github.io/Rmonize-documentation/reference/harmo_process.html), which uses 
information from the inputs jointly to generate harmonized datasets (one for 
each input dataset). 

```{r}

harmonized_dossier <- harmo_process(
    dossier, 
    dataschema, 
    data_proc_elem)

```

This produces a harmonized dossier (a list of harmonized datasets and associated 
metadata, with associated information from the DataSchema and DPE). If there 
were any errors or warnings during the process, these will be printed in the 
console. You can print a summary of any errors and warnings associated with 
individual algorithms in the console with 
[show_harmo_error()](https://maelstrom-research.github.io/Rmonize-documentation/reference/show_harmo_error.html).

```{r}

show_harmo_error(harmonized_dossier)

```


<em>
Note: If there is a processing error, a harmonized dossier will be created, 
but the affected harmonized dataset(s) will be empty. 
</em>

## Validate and consolidate harmonized data

You can assess and summarize harmonized datasets to identify potential issues, 
produce summary statistics, and generate visual reports. To perform evaluations
and summaries of the entire harmonized dossier, you can use 
[harmonized_dossier_evaluate()](https://maelstrom-research.github.io/Rmonize-documentation/reference/harmonized_dossier_evaluate.html) and
[harmonized_dossier_summarize()](https://maelstrom-research.github.io/Rmonize-documentation/reference/harmonized_dossier_summarize.html), 
which produce summary tables (that can be exported to Excel).

```{r}
# Evaluate and summarize a harmonized dossier containing multiple harmonized datasets.

harmonized_dossier_evaluation <- harmonized_dossier_evaluate(harmonized_dossier)
harmonized_dossier_summary <- harmonized_dossier_summarize(harmonized_dossier)

```

A visual report with summary figures for each variable can also be produced using [harmonized_dossier_visualize()](https://maelstrom-research.github.io/Rmonize-documentation/reference/harmonized_dossier_visualize.html). 

**Warning ⚠** *This tutorial creates a folder 'tmp' where the visual report is generated.*

```{r}
# place your harmonized dossier in a folder. This folder name is mandatory, and 
# must not previously exist.

bookdown_path <- paste0('tmp/',basename(tempdir()))

harmonized_dossier_visualize(harmonized_dossier, bookdown_path)

# Open the visual report in a browser.
bookdown_open(bookdown_path)

```

To combine the individual harmonized datasets in a harmonized dossier into one 
pooled harmonized dataset, use 
[pooled_harmonized_dataset_create()](https://maelstrom-research.github.io/Rmonize-documentation/reference/pooled_harmonized_dataset_create.html).

```{r}

# Generate one pooled harmonized dataset from a harmonized dossier
pooled_harmonized_dataset <- 
  pooled_harmonized_dataset_create(harmonized_dossier)

```


To get the harmonized data dictionary for an individual harmonized dataset 
(which contains more details about algorithms and R scripts used in processing 
for that dataset), use 
[data_dict_extract()](https://maelstrom-research.github.io/madshapR-documentation/reference/data_dict_extract.html).

```{r}

# Extract the harmonized data dictionary for one harmonized dataset.

harmonized_TOKYO_dd <- data_dict_extract(harmonized_dossier$dataset_TOKYO)

```


## Use harmonized data
Once you are satisfied with the outputs, they can be exported in any R 
compatible format. For example, harmonized datasets can be exported as 
labelled datasets that keep variable attributes as metadata (e.g. SAS files), 
and tabular reports can be exported as Excel files.

```{r}

library(fabR)
## Examples of exporting objects as Excel files.

# write_excel_allsheets(harmonized_dossier, "myfile.xlsx")
# write_excel_allsheets(harmonized_dossier_summary, "myfile.xlsx")
# write_excel_allsheets(harmonized_TOKYO_dd, "myfile.xlsx")

```


